DateLife: leveraging databases and analytical tools to reveal the dated Tree of Life

### 2022-05-05 manuscript/manuscript.Rmd
minor restcurture

DateLife processes input scientific names using a Taxonomic Name Resolution Service (TNRS), which increases the probability of correctly finding the queried taxon names in the chronogram database. TNRS detects, corrects and standardizes name misspellings and typos, variant spellings and authorities, and nomenclatural synonyms to a single taxonomic standard [@Boyle2013]. DateLife implements TNRS using OpenTree's unified taxonomy as standard [@opentreeAPIs; @rees2017automated], storing OpenTree's Taxonomy identification numbers for further processing.
<!-- We are using OpenTree's TNRS, Boyle described TNRS first, right?  -->

@@

1. The matching pruned chronograms as newick strings or "phylo" objects.
2. The ages of the root of all source chronograms. These ages can correspond to the age of the most recent common ancestor (mrca) of the user's group of interest if the source chronograms have all taxa belonging to the group. If not, the root corresponds to the mrca of a subgroup withing the group of interest.
3. Study citations where original chronograms were published.
4. A report of input taxon names matches across source chronograms.
5. The source chronogram(s) with the most input taxon names.
6. Various single summary chronograms resulting from summarizing age data, generated using the methodology described below.

@@

_**Dating the topology.--**_ 
Input topologies from OpenTree or the supertree approach described above do not include branch length estimates of any kind. 
To estimate branch lengths proportional to substitution rates on these topologies, DateLife can mine the Barcode of Life Data System, BOLD [@ratnasingham2007bold] to obtain genetic markers for the input taxa.
These markers are aligned with MUSCLE by default [@edgar2004muscle], or MAFFT [@katoh2009multiple]. 
This alignment can be used to estimate branch lengths on input topologies that lack branch lengths.
Currently, branch length reconstruction in DateLife is performed using parsimony and the likelihood of the phylogenetic tree given a sequence alignment is computed [@schliep2011phangorn].
While relative branch length information provides additional data for nodes without secondary date calibrations, topologies without branch lengths can also be dated.

@@

For each congruent node, the pairwise distance that traverse that node are summarized into a single summary matrix using classic summary statistics (i.e., mean, median, minimum and maximum ages), and the Supermatrix Distance Method [SDM; @Criscuolo2006], which deforms patristic distance matrices by minimizing variance and then averaging them.
These single summary taxon pair age matrices (Summarized calibrations) can be applied as secondary calibrations to date a tree topology, using different dating methods currently supported within DateLife: MrBayes [@Huelsenbeck2001; @Ronquist2003], PATHd8 [@Britton2007], BLADJ [@Webb2008; @webb2005phylomatic], and treePL [@Smith2012].

By default, DateLife implements the Branch Length Adjuster (BLADJ) algorithm to obtain a fully dated topology. BLADJ fixes node ages that have calibration data, and distributes time between nodes with no data evenly between nodes with calibration data.
This minimizes age variance in the resulting chronogram [@Webb2008].
BLADJ does not incorporate branch lengths even when they are present.
When there is conflict in ages between nodes with calibration data, the algorithm ignores ages that are older than ages of parent nodes and younger than ages from descendant nodes. 
> NOTE: do you know how it decides which one to drop if two calibrations conflict? The parent or the child? It would drop both. It basically drops any calibration for any given node that are older than any calibration of the parent node, and younger than any calibration of children nodes.

@@

Alternative options supported in DateLife (MrBayes, PATHD8, TreePL) incorporate branch length information from the input topology in combination with the secondary calibrations.
PATHd8 is a non-clock, rate-smoothing method [@Britton2007] to date trees.
treePL, is a semi-parametric, rate-smoothing, penalized likelihood dating method
[@Smith2012].
The MrBayes  [@Huelsenbeck2001; @Ronquist2003] approach in datelife uses the secondary calibrations as priors on node ages. 
> NOTE: is this true?? Yes

_**Visualizing results.--**_
Finally, users can save all source and summary chronograms in formats that permit reuse and reanalyses (newick and R "phylo" format), as well as visualize and compare results graphically, or construct their own graphs using \texttt{datelife}'s chronogram plot generation functions available from the R package `datelifeplot` [@datelifeplot].

>>>>

DateLife processes input scientific names using a Taxonomic Name Resolution Service (TNRS), which increases the probability of correctly finding the queried taxon names in the chronogram database. TNRS detects, corrects and standardizes name misspellings and typos, variant spellings and authorities, and nomenclatural synonyms to a single taxonomic standard [@boyle2013]. DateLife implements TNRS using OpenTree's unified taxonomy as standard [@opentreeAPIs; @rees2017automated], storing OpenTree's Taxonomy identification numbers for further processing.
<!-- We are using OpenTree's TNRS, Boyle described TNRS first, right?  -->

@@

1. The matching pruned chronograms as newick strings or "phylo" objects.
2. The ages of the root of all source chronograms. These ages can correspond to the age of the most recent common ancestor (mrca) of the user's group of interest if the source chronograms have all taxa belonging to the group. If not, the root corresponds to the mrca of a subgroup withing the group of interest.
3. Study citations where original chronograms were published.
4. A report of input taxon names matches across source chronograms.
5. The source chronogram(s) with the most input taxon names.
6. A single summary chronogram resulting from summarizing age data generated using the options described below.

@@

_**Dating the topology.--**_ 
Input topologies from OpenTree or the supertree approach described above do not include branch length estimates of any kind. 
To estimate branch lengths proportional to substitution rates on these topologies, DateLife can mine the Barcode of Life Data System, BOLD [@ratnasingham2007bold] to obtain genetic markers for the input taxa.
These markers are aligned with MUSCLE [@edgar2004muscle; or MAFFT, @katoh2009multiple]. 
This alignment can be used to estimate branch lengths on input topologies that lack branch lengths.
Currently, branch length reconstruction is performed with parsimony and the likelihood of the phylogenetic tree given a sequence alignment is computed [@schliep2011phangorn].

While relative branch length information provides additional data for nodes without secondary date calibrations, topologies without branch lengths can also be dated.

@@

<!-- we should calculate summary statistics per node -->
<!--??? How is 'calibrations' defined here? the calibrations from congruification, or the pairwise distances??? is calibrations from congruification-->
For each congruent node, the pairwise distance that traverse that node are summarized into a single summary matrix using classic statistics (mean, median, minimum and maximum ages). These single summary ages per taxon pair are then used as secondary calibrations to date the tree topology.  
<!--
Under the SDM criterion ??? what happens??
Summarizing options implemented include Super Distance Matrix method [SDM, @Criscuolo2006] 
-->
Summarized calibrations can be applied as secondary calibrations with different dating methods currently supported within DateLife: MrBayes [@Huelsenbeck2001; @Ronquist2003], PATHd8 [@Britton2007], BLADJ [@Webb2008; @webb2005phylomatic], and treePL [@Smith2012].

By default, DateLife implements the Branch Length Adjuster (BLADJ) algorithm to obtain a fully dated topology. BLADJ fixes node ages that have calibration data, and distributes time between nodes with no data evenly between nodes with calibration data.
This minimizes age variance in the resulting chronogram [@Webb2008].
BLADJ does not incorporate branch lengths even when they are present.
When there is conflict in ages between nodes with calibration data, the algorithm ignores ages that are older than ages of parent nodes and/or younger than ages from descendant nodes. 
\NOTE:do you know how it decides which one to drop if two calibrations conflict? The parent or the child?

@@

Alternative options supported in DateLife (MrBayes, PATHD8, TreePL) incorporate branch length information from the input topology in combination with the secondary calibrations.
PATHd8 is a non-clock, rate-smoothing method [@Britton2007] to date trees.
treePL, is a semi-parametric, rate-smoothing, penalized likelihood dating method
[@Smith2012].
The MrBayes  [@Huelsenbeck2001; @Ronquist2003] apprach in datelife uses the secondary calibrartions as priors on node ages. \NOTE: is this true??

_**Visualizing results.--**_
Finally, users can save all source and summary chronograms in formats that permit reuse and reanalyses (newick and R "phylo" format), as well as visualize and compare results graphically, or construct their own graphs using \texttt{datelife}'s chronogram plot generation functions.
